---
title: "VTPEH 6270 Checkpoint 6"
author: "Sowmya Srinivasan"
date: "`r Sys.Date()`"
editor_options:
  chunk_output_type: console
output:
  pdf_document:
    toc: true
    toc_depth: 1
    number_sections: true
    latex_engine: pdflatex
header-includes:
  - \usepackage{float}
urlcolor: blue
---

```{r setup, include = FALSE}

knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE,
                      fig.align = "center", out.width = "85%",
                      fig.pos = "H")

```

# Introduction and Research Question

This analysis uses the COVID-19 Vaccine Hesitancy for County and Local Estimates data set published by the U.S. Centers for Disease Control and Prevention (CDC). The dataset contains county-level estimates of vaccine hesitancy among adults, vaccination coverage as of June 10, 2021, and measures of social vulnerability, including the Social Vulnerability Index (SVI) and demographic characteristics. Each observation represents a U.S. county and includes both continuous and categorical variables suitable for exploratory data analysis and visualization.
The complete data set and accompanying information are available at 
https://data.cdc.gov/Vaccinations/Vaccine-Hesitancy-for-COVID-19-County-and-local-es/q9mh-h2tw/about_data


Research Question:

Is COVID-19 vaccine hesitancy higher in counties with higher Social Vulnerability Index (SVI)

Variable A: Estimated hesitant - percentage of adults estimated to be hesitant toward COVID-19 vaccination; continuous variable

Variable B: SVI category - categorized measure of county-level social vulnerability; categorical variable

# Material & Methods

## Loading Data and Data Description

### Loading Data

```{r libraries}
library(readxl)
library(dplyr)
library(ggplot2)
library(knitr)
library(kableExtra)
library(tidyr)


```

```{r load-data, results='hide'}

# Set working directory
setwd("/Users/sowmya427/Desktop/VTPEH-6270-R-Lab/data")

# Load file
raw <- read.csv("R Lab_Vaccine_Hesitancy.csv", check.names = FALSE)

# Confirm columns loaded correctly
cat("Column names:\n")
print(names(raw))

df <- raw %>%
  rename(
    fips               = `FIPS Code`,
    county             = `County Name`,
    state              = `State`,
    hesitant           = `Estimated hesitant`,
    hesitant_or_unsure = `Estimated hesitant or unsure`,
    strongly_hesitant  = `Estimated strongly hesitant`,
    svi                = `Social Vulnerability Index (SVI)`,
    svi_cat            = `SVI Category`,
    vax_rate           = `Percent adults fully vaccinated against COVID-19 (as of 6/10/21)`,
    pct_hispanic       = `Percent Hispanic`,
    pct_black          = `Percent non-Hispanic Black`,
    pct_white          = `Percent non-Hispanic White`
  ) %>%
  mutate(
    hesitant_pct = as.numeric(gsub("%", "", hesitant)),
    vax_rate_pct = as.numeric(gsub("%", "", vax_rate)),
    svi_cat = factor(
      svi_cat,
      levels = c(
        "Very Low Vulnerability",
        "Low Vulnerability",
        "Moderate Vulnerability",
        "High Vulnerability",
        "Very High Vulnerability"
      )
    )
  ) %>%
  filter(!is.na(svi), !is.na(hesitant_pct))

cat("Final sample size:", nrow(df), "counties\n")
```


### Nature Theme 

```{r nature-theme}
dir.create("figures/CP06", recursive = TRUE, showWarnings = FALSE)

theme_nature <- function(base_size = 10) {
  theme_classic(base_size = base_size) +
    theme(
      text             = element_text(family = "sans"),
      axis.title       = element_text(size = base_size, color = "black"),
      axis.text        = element_text(size = base_size - 2, color = "black"),
      axis.line        = element_line(color = "black", linewidth = 0.5),
      axis.ticks       = element_line(color = "black", linewidth = 0.5),
      legend.text      = element_text(size = base_size - 2),
      legend.title     = element_text(size = base_size - 1, face = "bold"),
      legend.key.size  = unit(0.4, "cm"),
      plot.title       = element_text(size = base_size + 1, face = "bold", hjust = 0),
      plot.caption     = element_text(size = base_size - 3, color = "gray50", hjust = 0),
      panel.border     = element_blank(),
      strip.background = element_blank()
    )
}
```


### Description Table

```{r summary-table}
summary_tbl <- df %>%
  group_by(svi_cat) %>%
  summarise(
    n      = n(),
    Mean   = round(mean(hesitant_pct), 2),
    SD     = round(sd(hesitant_pct),   2),
    Median = round(median(hesitant_pct), 2),
    Min    = round(min(hesitant_pct),  2),
    Max    = round(max(hesitant_pct),  2),
    .groups = "drop"
  )

kable(
  summary_tbl,
  col.names = c("SVI Category", "n", "Mean (\\%)", "SD",
                "Median (\\%)", "Min (\\%)", "Max (\\%)"),
  caption  = "Descriptive statistics of estimated COVID-19 vaccine hesitancy (\\%) by SVI category.",
  booktabs = TRUE
) %>%
  kable_styling(
    latex_options = c("hold_position", "striped"),
    font_size     = 10
  )

```


## Statistical Analysis

To examine whether COVID-19 vaccine hesitancy rates differ across Social 
Vulnerability Index (SVI) categories, the following statistical approach 
was used.

**Assumption Checks**

Prior to selecting a statistical test, the normality of the outcome 
variable (estimated vaccine hesitancy rate, %) was assessed using the 
Shapiro-Wilk test, where a p-value < 0.05 indicates a significant 
departure from normality. A histogram and Q-Q plot were also used to 
visually inspect the distribution of the data. Normality is a required 
assumption for parametric tests such as one-way ANOVA, and failure to 
meet this assumption requires the use of non-parametric alternatives.

**Model Selection**

Because the Shapiro-Wilk test confirmed that the hesitancy rate data 
were non-normally distributed, a Kruskal-Wallis rank-sum test was 
selected to compare hesitancy rates across the five SVI categories 
(Very Low, Low, Moderate, High, and Very High Vulnerability). The 
Kruskal-Wallis test is the non-parametric equivalent of a one-way ANOVA 
and does not require the assumption of normality. In addition, 
Spearman's rank correlation was used to quantify the strength and 
direction of the monotonic association between the continuous SVI score 
and vaccine hesitancy rate. Spearman's correlation was chosen over 
Pearson's correlation as it does not assume normality and is appropriate 
for non-normally distributed continuous variables.

**Post-Hoc Analysis**

A significant Kruskal-Wallis result indicates that at least one SVI 
category differs from the others, but does not specify which pairs. 
Therefore, pairwise Wilcoxon rank-sum tests were conducted as a 
post-hoc analysis to identify which specific SVI category pairs showed 
statistically significant differences in vaccine hesitancy rates. 
P-values were adjusted using the Bonferroni correction to control for 
multiple comparisons and reduce the risk of Type I error. Statistical 
significance was set at alpha = 0.05 for all analyses.

## Code

All code and data used in this analysis are available at: https://github.com/SS0908/VTPEH-6270-R-Lab


# Results

## Data Visualization 


### Boxplot
```{r boxplot, fig.cap="Distribution of estimated COVID-19 vaccine hesitancy (\\%) by SVI category. Boxes span the interquartile range; horizontal lines show medians; dots are outliers."}
p_box <- ggplot(df, aes(x = svi_cat, y = hesitant_pct, fill = svi_cat)) +
  geom_boxplot(
    alpha         = 0.80,
    outlier.size  = 0.6,
    outlier.alpha = 0.35,
    linewidth     = 0.4
  ) +
  scale_fill_manual(
    name   = "SVI Category",
    values = c(
      "Very Low Vulnerability"  = "#4393C3",
      "Low Vulnerability"       = "#92C5DE",
      "Moderate Vulnerability"  = "#D9D9D9",
      "High Vulnerability"      = "#F4A582",
      "Very High Vulnerability" = "#D6604D"
    )
  ) +
  scale_x_discrete(
    labels = c("Very Low", "Low", "Moderate", "High", "Very High")
  ) +
  labs(
    title   = "COVID-19 Vaccine Hesitancy by Social Vulnerability Category",
    x       = "SVI Category",
    y       = "Estimated Hesitancy (%)",
    caption = "Source: CDC Vaccine Hesitancy County Estimates"
  ) +
  theme_nature()

p_box
```

```{r boxplot-save, include=FALSE}
ggsave("figures/CP06/CP06_boxplot.pdf",
       plot = p_box, width = 7, height = 4.5, device = "pdf")
```



### Scatterplot

```{r scatter, fig.cap="Association between continuous SVI score and estimated vaccine hesitancy (\\%) across U.S. counties (n = 3,141). Each point is one county. Red curve is a LOESS smoother with 95\\% CI shaded."}
p_scatter <- ggplot(df, aes(x = svi, y = hesitant_pct)) +
  geom_point(
    aes(color = "County data"),
    alpha = 0.15,
    size  = 0.6
  ) +
  geom_smooth(
    aes(color = "LOESS smoother"),
    method    = "loess",
    linewidth = 1,
    se        = TRUE,
    fill      = "#F4A582",
    alpha     = 0.3
  ) +
  scale_color_manual(
    name   = "Legend",
    values = c(
      "County data"    = "#4393C3",
      "LOESS smoother" = "#D6604D"
    )
  ) +
  labs(
    title   = "Social Vulnerability Index vs. COVID-19 Vaccine Hesitancy",
    x       = "Social Vulnerability Index (SVI, 0-1)",
    y       = "Estimated Hesitancy (%)",
    caption = "Source: CDC Vaccine Hesitancy County Estimates"
  ) +
  theme_nature()

p_scatter
```

```{r scatter-save, include=FALSE}
ggsave("figures/CP06/CP06_scatterplot.pdf",
       plot = p_scatter, width = 7, height = 4.5, device = "pdf")
```







### Normality; Plot 1 - Histogram

```{r normality-hist, fig.cap="Histogram of estimated COVID-19 vaccine hesitancy rates (\\%) across U.S. counties. Dashed red line = mean; dotted green line = median.", fig.height=4}
p_hist <- ggplot(df, aes(x = hesitant_pct)) +
  geom_histogram(
    aes(fill = "Counties (n = 3,141)"),
    bins  = 40,
    color = "white",
    alpha = 0.85
  ) +
  geom_vline(
    aes(xintercept = mean(hesitant_pct, na.rm = TRUE), color = "Mean"),
    linewidth = 1,
    linetype  = "dashed"
  ) +
  geom_vline(
    aes(xintercept = median(hesitant_pct, na.rm = TRUE), color = "Median"),
    linewidth = 1,
    linetype  = "dotted"
  ) +
  scale_fill_manual(
    name   = "Distribution",
    values = c("Counties (n = 3,141)" = "#4393C3")
  ) +
  scale_color_manual(
    name   = "Reference Lines",
    values = c("Mean" = "#D6604D", "Median" = "#4DAF4A")
  ) +
  labs(
    title = "Distribution of COVID-19 Vaccine Hesitancy Rates",
    x     = "Estimated Hesitancy (%)",
    y     = "Number of Counties"
  ) +
  theme_nature() +
  theme(legend.position = "right")

p_hist
```

```{r histogram-save, include=FALSE}
ggsave("figures/CP06/CP06_histogram.pdf",
       plot = p_hist, width = 7, height = 4, device = "pdf")
```







### Normality; Plot 2 - Q-Q plot

```{r normality-qq, fig.cap="Q-Q plot of estimated vaccine hesitancy rates. Blue points = sample data; red line = expected values under a normal distribution. Deviations at the tails indicate non-normality."}
p_qq <- ggplot(df, aes(sample = hesitant_pct)) +
  stat_qq(
    aes(color = "Sample data"),
    alpha = 0.3,
    size  = 0.8
  ) +
  stat_qq_line(
    aes(color = "Normal reference line"),
    linewidth = 1
  ) +
  scale_color_manual(
    name   = "Legend",
    values = c(
      "Sample data"           = "#4393C3",
      "Normal reference line" = "#D6604D"
    )
  ) +
  labs(
    title = "Q-Q Plot of Vaccine Hesitancy Rates",
    x     = "Theoretical Quantiles",
    y     = "Sample Quantiles (%)"
  ) +
  theme_nature() +
  theme(legend.position = "bottom")

p_qq
```


```{r qqplot-save, include=FALSE}
ggsave("figures/CP06/CP06_qqplot.pdf",
       plot = p_qq, width = 6, height = 5, device = "pdf")
```





## Statistical Analysis

### Shapiro-Wilk Test

```{r shapiro-test}
sw <- shapiro.test(df$hesitant_pct)

sw_tbl <- data.frame(
  Test        = "Shapiro-Wilk",
  W           = round(sw$statistic, 4),
  `p-value`   = ifelse(sw$p.value < 0.001, "< 0.001", round(sw$p.value, 4)),
  Result      = ifelse(sw$p.value < 0.05,
                       "Non-normal - use non-parametric tests",
                       "Normal distribution"),
  check.names = FALSE
)

kable(
  sw_tbl,
  caption  = "Shapiro-Wilk normality test for vaccine hesitancy rates.",
  booktabs = TRUE,
  align    = c("l", "c", "c", "l")
) %>%
  kable_styling(
    latex_options = c("hold_position", "striped"),
    font_size     = 10
  )
```


The Shapiro-Wilk test returned a W statistic of `r round(sw$statistic, 4)` 
and a p-value < 0.001, indicating that the distribution of vaccine hesitancy 
rates departs significantly from normality. This means we cannot assume the 
data follow a normal distribution, and therefore non-parametric tests must be 
used for all subsequent group comparisons rather than parametric alternatives 
such as ANOVA or a t-test.

### Kruskal-Walis Test

```{r kruskal}
kw <- kruskal.test(hesitant_pct ~ svi_cat, data = df)

kw_tbl <- data.frame(
  Test          = "Kruskal-Wallis Rank-Sum Test",
  `Chi-squared` = round(kw$statistic, 3),
  df            = kw$parameter,
  `p-value`     = ifelse(kw$p.value < 0.001, "< 0.001", round(kw$p.value, 4)),
  Result        = "Significant difference in hesitancy across SVI categories",
  check.names   = FALSE
)
rownames(kw_tbl) <- NULL

kable(
  kw_tbl,
  caption  = "Kruskal-Wallis test comparing vaccine hesitancy rates across SVI categories.",
  booktabs = TRUE,
  align    = c("l", "c", "c", "c", "l")
) %>%
  kable_styling(
    latex_options = c("hold_position", "striped", "scale_down"),
    font_size     = 10
  ) %>%
  column_spec(5, width = "5cm")
```

The Kruskal-Wallis rank-sum test revealed a statistically significant 
difference in vaccine hesitancy rates across the five SVI categories 
(X^2 = `r round(kw$statistic, 2)`, df = `r kw$parameter`, p < 0.001). 
This result allows us to reject the null hypothesis that hesitancy rates 
are equal across SVI categories. In other words, social vulnerability level 
is associated with differences in COVID-19 vaccine hesitancy at the county 
level. However, this test alone does not tell us which specific categories 
differ from one another, which is addressed by the pairwise comparisons below.


### Pairwise Wilcoxon 

```{r pairwise-wilcox}
pw <- pairwise.wilcox.test(
  x               = df$hesitant_pct,
  g               = df$svi_cat,
  p.adjust.method = "bonferroni"
)

# Format the p-value matrix into a clean table
pw_tbl <- as.data.frame(pw$p.value) %>%
  tibble::rownames_to_column(var = "Group") %>%
  mutate(across(where(is.numeric), ~ case_when(
    . < 0.001 ~ "< 0.001",
    . < 0.05  ~ as.character(round(., 3)),
    is.na(.)  ~ "--",
    TRUE      ~ as.character(round(., 3))
  )))

kable(
  pw_tbl,
  col.names = c("Group", "Very Low", "Low", "Moderate", "High"),
  caption  = "Pairwise Wilcoxon rank-sum tests comparing vaccine hesitancy across SVI categories (Bonferroni correction). '--' = comparison not applicable.",
  booktabs = TRUE,
  align    = "c"
) %>%
  kable_styling(
    latex_options = c("hold_position", "scale_down", "striped"),
    font_size     = 9
  ) %>%
  column_spec(1, width = "4cm")
```


Pairwise Wilcoxon rank-sum tests with Bonferroni correction were conducted 
to identify which specific SVI category pairs drive the overall significant 
Kruskal-Wallis result. The results show that most pairwise comparisons 
between SVI categories were statistically significant (p < 0.001) after 
correction for multiple testing. Counties in higher vulnerability categories 
consistently showed greater vaccine hesitancy than those in lower 
vulnerability categories. Comparisons between adjacent categories (e.g., 
Very Low vs. Low Vulnerability) may show weaker or non-significant 
differences, reflecting the gradual nature of the relationship between 
social vulnerability and hesitancy.


### Spearman Correlation
```{r spearman}
sp <- cor.test(df$svi, df$hesitant_pct, method = "spearman")

sp_tbl <- data.frame(
  Test        = "Spearman Rank Correlation",
  `rho (p)`   = round(sp$estimate, 3),
  S           = round(sp$statistic, 1),
  `p-value`   = ifelse(sp$p.value < 0.001, "< 0.001", round(sp$p.value, 4)),
  Direction   = "Positive association",
  Result      = "Higher SVI linked to higher hesitancy",
  check.names = FALSE
)
rownames(sp_tbl) <- NULL

kable(
  sp_tbl,
  caption  = "Spearman rank correlation between continuous SVI score and vaccine hesitancy (\\%).",
  booktabs = TRUE,
  align    = c("l", "c", "c", "c", "l", "l")
) %>%
  kable_styling(
    latex_options = c("hold_position", "scale_down", "striped"),
    font_size     = 10
  ) %>%
  column_spec(1, width = "3.5cm") %>%
  column_spec(5, width = "2.5cm") %>%
  column_spec(6, width = "3cm")
```


Spearman's rank correlation between the continuous SVI score and vaccine 
hesitancy rate yielded a correlation coefficient of 
p = `r round(sp$estimate, 3)` (p < 0.001), indicating a statistically 
significant positive monotonic association. This means that as the Social 
Vulnerability Index increases across counties, estimated vaccine hesitancy 
also tends to increase. The magnitude of the correlation suggests a moderate 
positive relationship, supporting the conclusion that social vulnerability 
is meaningfully — though not perfectly — linked to COVID-19 vaccine 
hesitancy at the county level. This finding directly addresses our research 
question and is consistent with the patterns observed visually in the 
scatterplot and boxplot.


# Conclusion


The results of this analysis support the hypothesis that COVID-19 vaccine 
hesitancy is higher in counties with greater social vulnerability. The 
Kruskal-Wallis test identified statistically significant differences in 
hesitancy rates across SVI categories (X^2 = `r round(kw$statistic, 2)`, 
p < 0.001), and pairwise Wilcoxon comparisons confirmed that 
higher-vulnerability counties showed greater hesitancy than 
lower-vulnerability counties across most category pairs. The Spearman 
correlation (p = `r round(sp$estimate, 3)`, p < 0.001) further supports 
a moderate positive association between the continuous SVI score and 
estimated hesitancy rate.

These findings align with the research question posed in the Introduction 
— counties with higher social vulnerability do tend to have higher 
estimated COVID-19 vaccine hesitancy. The moderate strength of the 
association suggests that while social vulnerability is a meaningful 
predictor of hesitancy, other factors likely contribute as well. These 
results highlight the importance of targeting socially vulnerable 
communities in public health vaccination campaigns.


# AI Use Disclosure Statement

This document was created with the assistance of Claude (Anthropic) to help structure and generate the R code, which was reviewed and adjusted by the author. Interpretations and written sections were composed by the author.
